---
title: "Analyzing Voting Difficulty"
subtitle: "Datasci 203: Lab 1 Part 2"
author: "Zachary Fenton, Rob Horrobin, Yoni Nackash" 
output:
  bookdown::pdf_document2: 
    toc: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{=tex}
\newpage
\setcounter{page}{1}
```
```{r load packages and set options, include=FALSE, echo=FALSE}
library(tidyverse) 
library(magrittr)
library(knitr)
library(patchwork)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(kableExtra)

theme_set(theme_bw())

options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
```

```{r, include=FALSE, echo=FALSE}
# Load ANES Pilot Study 2022 data
anes <- read.csv('~/lab-1-catch-22/datasets/anes_pilot_2022_csv_20221214.csv')

# Create a smaller dataframe with key variables of interest
simple_anes<-anes %>%
  select(weight, pid_x, pid1d, pid1r, pidstr, pidlean, turnout22, pipevote22a, turnout22ns, turnout20, votehard, regdiff)

# Creating new column "Affiliation" that identifies a respondents political party
simple_anes$Affiliation=''
simple_anes$Affiliation = ifelse(simple_anes$pid_x >=5, "Republican",ifelse(simple_anes$pid_x <=3,"Democrat", "Independent"))

# Creating new column "Strength" that identifies a respondents strength of affiliation to a political party
#Helpful reference https://stackoverflow.com/questions/11865195/using-if-else-on-a-data-frame
simple_anes$Strength=''
simple_anes$Strength= ifelse(simple_anes$pid_x == 1 | simple_anes$pid_x == 7, "Strong",ifelse(simple_anes$pid_x == 2 | simple_anes$pid_x == 6,"Weak", ifelse(simple_anes$pid_x == 3 | simple_anes$pid_x == 5,"Lean", "Independent")))

# Define cohort of voters, and create new column "voter"
# Voters reponded 1,2,3 to "turnout22", 2 to "pipevote22a" or 1 to "turnout22ns"
simple_anes$voter <- 0
simple_anes$voter[simple_anes$turnout22==1 | simple_anes$turnout22==2 | simple_anes$turnout22==3 | simple_anes$pipevote22a==2 | simple_anes$turnout22ns==1] <- 1

# Remove Independent voters, and voters who did not respond to "votehard" question
# Create our analytics base table for model ingestion
abt<- simple_anes %>%
  filter(Affiliation != "Independent" & voter !=0  & votehard !=-1)


# Creating inline code chunks for write-up
nrow_original = nrow(simple_anes) # 1,585

nrow_dem_rep = length(which(simple_anes$Affiliation !='Independent')) #1,310

nrow_dem_rep_turnout22 = length(which(simple_anes$Affiliation !='Independent' & 
               (simple_anes$turnout22==1 | simple_anes$turnout22==2 | simple_anes$turnout22==3))) # 726

nrow_dem_rep_pipevote22a = length(which(simple_anes$Affiliation !='Independent' 
                                        & simple_anes$pipevote22a==2 
                                        & (simple_anes$turnout22!=1 | simple_anes$turnout22!=2 | simple_anes$turnout22!=3))) # 318

nrow_dem_rep_turnout22ns = length(which(simple_anes$Affiliation !='Independent' 
                                        & simple_anes$turnout22ns==1 
                                        & simple_anes$pipevote22a!=2
                                        & (simple_anes$turnout22!=1 | simple_anes$turnout22!=2 | simple_anes$turnout22!=3))) # 3

nrow_dem_rep_voters = length(which(simple_anes$Affiliation !='Independent' 
                                   & simple_anes$voter=='1'
                                   & simple_anes$votehard!=-1)) # 1,047
```

# Importance and Context

Voting is an extremely vital component of a healthy democracy. A
substantial electorate ensures that as many citizens' opinions and
voices are represented in the government as possible. As the United
States is becoming increasingly more aware of systemic biases within our
country, the topic of success versus struggle when voting is an
important and interesting topic to be analyzed. There may in fact be
inherent biases in our voting system that cause more difficulty to
certain kinds of voters than others. We are interested in answering the
following question:

> *Do Democratic voters or Republican voters experience more difficulty
> voting?*

Answering this question will help provide insight into potential
systemic biases in the structure and operations of elections in the
United States. These insights will be key for both local governments as
well as the federal government to make positive changes to our voting
process, and reduce the difficulty voters experience.

# Data and Methodology

We used data from the American National Election Studies (ANES)[^1] to
study the challenges faced by Democratic and Republican voters and how
these experiences may vary based on political party preference. This
dataset has been collected since 1948, and we focused on the 2022 Pilot
Study, which examines voting and public opinion after the 2022 midterm
elections in the United States.

[^1]: American National Election Studies, 2022, "2022 Pilot Study",
    <https://electionstudies.org/data-center/2022-pilot-study/>

The 2022 Pilot Study contains `r nrow_original` cases, which are a subset of over a
million volunteers who responded to an internet survey on YouGov for a
small compensation. These `r nrow_original` cases were selected by study designers
to match the overall U.S. population in terms of gender, age, race, and
education. While study designers calculated weightings to represent
other characteristics, for this analysis, we didn't consider these
weightings.

We filtered our data based on responses to several survey questions:

(1) Affiliation: respondents with strong, weak or leaning preference to
    the Democrat or Republican parties.
(2) Voter Identification: based on survey responses, who do we consider
    a voter?
(3) Voting Difficulty: evaluating the level of difficulty voters
    experienced in the 2022 election.

To meet the first criterion, we analyzed survey responses where
respondents claimed to identify as a Democrat, Republican, Independent,
or other. We also considered the strength of their affiliation, ranging
from *Lean* to *Weak* to *Strong* as self-reported. Research by Petrocik
(2009)[^2] suggested that lean or weak persuasions indicate a clear
preference for a particular party, so we treated *Lean* and *Weak*
preferences as a reliable indicator of political preference. Considering
our question focuses on Democrats and Republicans, we removed
respondents who identify as Independents, other, or NA. Out of the `r nrow_original`
cases, this results in `r nrow_dem_rep` respondents identifying as either Democrat
or Republican. Reference Table 1 for a summary of political affiliation.

[^2]: John Richard Petrocik, (2009, July 2). Measuring party support:
    Leaners are not independents. Electoral Studies.
    <https://www.sciencedirect.com/science/article/abs/pii/S0261379409000511>

```{r, message=FALSE, echo=FALSE, fig.align='center', fig.show='hold', fig.cap='Self reported political identification'}
# Create political party summary table, table 1
strong_dem = nrow(filter(simple_anes, Affiliation == "Democrat", Strength == "Strong"))
weak_dem = nrow(filter(simple_anes, Affiliation == "Democrat", Strength == "Weak"))
lean_dem = nrow(filter(simple_anes, Affiliation == "Democrat", Strength == "Lean"))
lean_rep = nrow(filter(simple_anes, Affiliation == "Republican", Strength == "Lean"))
weak_rep = nrow(filter(simple_anes, Affiliation == "Republican", Strength == "Weak"))
strong_rep = nrow(filter(simple_anes, Affiliation == "Republican", Strength == "Strong"))

type <- c("Strong Democrat", "Weak Democrat", "Lean Democrat" , "Lean Republican", "Weak Republican", "Strong Republican")
values <- c(strong_dem, weak_dem, lean_dem, lean_rep, weak_rep, strong_rep)

party_summary <- data.frame(type, values)
colnames(party_summary) <- c('Self Identified Party','Number of Respondents') 

party_summary%>%
  kbl(caption = "Political affiliation vs strength", booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")
```

Second, we are interested in analyzing respondents that voted in 2022.
To identify these respondents, we included those who claimed to have
either definitely voted, most likely voted, and probably voted in 2022.
Out of the `r nrow_dem_rep` respondents that identify as either Democrat or
Republican, `r nrow_dem_rep_turnout22` respondents claimed to have definitely voted, `r nrow_dem_rep_pipevote22a`
respondents claimed to have most likely voted, and `r nrow_dem_rep_turnout22ns` respondents claimed
to have probably voted. This leaves us with a sample size of `r nrow_dem_rep_voters`
voters.

Third, we determined a parameter to measure difficulty. We opted to
focus on the general question *"How difficult was it for you to vote?"*
as our target variable, rather than attempt to combine various questions
of difficulty into a new parameter. We believe a combination of several
questions could have reduced practical significance, as we would be
unable to deconvolute the new scale of this parameter. The respondents
answered the general question of difficulty on Likert scale from 1 - 5,
where 1 corresponds to *Not difficult at all* and 5 corresponds to
*Extremely difficult*. These responses are subjective and, like the rest
of the ANES data is self-reported without direct measurement. Thus, we
must evaluate our results in the context that perception of respondents
may play a factor and may not be objectively conclusive.

As seen in Figure 1 below, the majority of voters in both parties
identified with ranking 1, claiming to experience very little difficulty
in casting their vote.

```{r, message=FALSE, echo=FALSE, fig.height = 3, fig.width = 10, fig.align='center', fig.show='hold', fig.cap='Distribution of voting difficulty by party'}
# Create Difficulty Distribution plots, figure 1
dem_abt <- abt[abt$Affiliation == 'Democrat',]
rep_abt <- abt[abt$Affiliation == 'Republican',]

gg1 <- ggplot(dem_abt, aes(x = votehard)) +
  geom_bar(width = 0.4, fill='dodgerblue1') +
  ggtitle('Democratic Voters') +
  #theme(text = element_text(size=17))
  theme(axis.title.x=element_blank(), 
        axis.title.y=element_blank(),
        plot.title = element_text(size=16, face="bold", hjust=0.5, vjust = 0.5))

gg2 <- ggplot(rep_abt, aes(x = votehard)) +
  geom_bar(width = 0.4, fill='firebrick1') +
  ggtitle('Republican Voters') +
  theme(axis.title.x=element_blank(), 
        axis.title.y=element_blank(),
        plot.title = element_text(size=16, face="bold", hjust=0.5, vjust = 0.5))

grid.arrange(gg1, gg2,nrow=1, left='Count of Voters', bottom = 'Difficulty Ranking') 
```

As stated above, we have identified two random variables - the
difficulty in voting on a Likert scale, for both Democrats and
Republicans. Because these data are ordinal and non-parametric, we
selected the Wilcoxon Rank-Sum Hypothesis of Comparisons test over other
two sample tests, such as the Two-Sample T-Test. We also acknowledge
that based on the large sample size (*n=`r nrow_dem_rep_voters`*), the central limit
theorem would suggest that the sampling distribution of the statistic
should be approximately normal, which would allow us to use a parametric
test. We decided against this route due to the data being ordinal.

The null hypothesis of our Wilcoxon Rank-Sum Test is:

```{=tex}
\begin{quote}
  \textbf{Null Hypothesis:} \textit{The probability that a Democrat identifying voter finds it difficult to vote is equal to the probability that a Republican identifying voter finds it difficult to vote.}
\end{quote}
```
For the Wilcoxon Rank-Sum test to be a valid test, the following must be
true, or assumed true:

First, the variables being measured are on an ordinal scale. As
indicated above, our target parameter answers the question *"How
difficult was it for you to vote?"* with answers ranging on a Likert
scale of *Not difficult at all* to *Extremely difficult*. This scale is
ordinal due to the ranking of the values, as well as the difference
between each value being subjective.

Second, the data is drawn from an I.I.D. sample. Voters from the same
party could experience similar difficulties in voting based on a number
of considerations, such as socioeconomic status or distance to polling
location. We acknowledge that this could introduce dependencies within
our samples. However, given that sample is drawn from a population of
millions of YouGov respondents, this would suggest that dependency or
links between individuals would be uncommon. Thus, for the purposes of
this analysis, we assume I.I.D is satisfied.

Third, the shapes of the distributions are roughly the same. We can see
in Figure 1 that this is reasonably satisfied.

# Results

```{r echo=TRUE}
test <- wilcox.test(votehard ~ Affiliation, data = abt, exact = FALSE)
```

```{r, message=FALSE, echo=FALSE}
# Creating inline code chunks for write-up
p_value = test$p.value
```

The final result of the Wilcoxon Rank-Sum test provides sufficient
evidence to reject the null hypothesis, (p=`r p_value`). This means that
there is a perceived difference in the way Democratic voters experience
difficulty versus Republican voters. However, through our two-tailed
test alone we are unable interpret which party may find it more
difficult. To help inform this, we standardized each sample distribution
to 1, allowing us to compare distribution of difficulty across parties
irrespective of population size. In Figure 2 it becomes clear that while
the majority of both Democrats and Republicans responded to feeling no
difficulty, there is a slightly larger distribution of Democrats
responding 2 *A little difficult* and 3 *Moderately difficult* than
Republicans.

```{r, message=FALSE, echo=FALSE, fig.height = 4, fig.width = 10, fig.align='center', fig.show='hold', fig.cap='Comparing difficulty by party, as a percentage'}
#Create standardized Democrat vs Republican stacked bar plot, figure 2
tab_dif_percent <- as.table(rbind(c(82.5, 91.9), c(10.5, 5.4), c(5.7, 2.1), c(1.2, 0.2), c(0.4, 0.4)))
dimnames(tab_dif_percent) <- list(Difficulty = c(1, 2, 3, 4, 5),
                                  Party = c("Democrat", "Republican"))

mx_df <- as.data.frame.matrix(tab_dif_percent)
data_long <- tidyr::gather(mx_df, key=type_col, value=categories)
data_long <- cbind(Difficulty=c(1,2,3,4,5,1,2,3,4,5), data_long)

stacked_plot <- ggplot(data_long, aes(x = Difficulty, y=categories, fill = type_col)) +
  geom_bar(stat = 'identity', position='dodge') +
  scale_fill_manual("Legend", values=c("Democrat" = "dodgerblue1", "Republican"="firebrick1")) +
  xlab("Difficulty Ranking") +
  ylab("Percent (%)")+
  ggtitle("Democrat vs Republican Voting Difficulty Comparison") +
  theme(legend.position = c(0.8,.8),
        plot.title = element_text(size=16, face="bold", hjust=0.5, vjust = 0.5),
        axis.title.x=element_text(size=14), 
        axis.title.y=element_text(size=14))

stacked_plot
```

# Discussion

```{r, include=FALSE, echo=FALSE}
# Calculate percent of Democrats and Republicans that experience difficulty
percent_dems_difficult = length(which(abt$Affiliation=='Democrat' & abt$votehard != '1')) / length(which(abt$Affiliation=='Democrat'))
percent_reps_difficult = length(which(abt$Affiliation=='Republican' & abt$votehard != '1')) / length(which(abt$Affiliation=='Republican'))

# Read in 2022 General Population voting data and calculate number of democratic votes
gen_election_22 <- read.csv('~/lab-1-catch-22/datasets/1976-2022-house.csv')
total_dem_votes_22 = sum(gen_election_22$candidatevotes[gen_election_22$year=='2022' & gen_election_22$party=='DEMOCRAT']) # 51808690

# Using percent of Democrats that experience difficulty from survey, extrapolate to entire US population
percent_dems_difficult * total_dem_votes_22 # 9022339

# Creating inline code chunks for write-up
percent_dems_difficult_chunk = round(percent_dems_difficult * 100,2)
percent_reps_difficult_chunk = round(percent_reps_difficult * 100,2)
total_dem_votes_22_chunk = round(total_dem_votes_22/1000000)
```

Our study has shown evidence that Democrats experience more difficulty
voting than Republicans. While our results do not indicate direct
causation between affiliated party and difficulty in voting, there is a
perceived relationship between a voter's political identification and
how much difficulty they experience while voting. We have found that
`r percent_dems_difficult_chunk` percent of Democratic voters experience difficulty voting, while `r percent_reps_difficult_chunk` percent of
Republican voters experience difficulty. The effect appears to be
practically significant, with Democrats twice as likely to experience
difficulty than Republicans. Given that the ANES 2022 Pilot Study is
weighted and intended to represent the entire US population, `r percent_dems_difficult_chunk` percent of
the roughly `r total_dem_votes_22_chunk` million Democratic voters from the 2022 General
Election[^3] could represent millions of voters. This is a considerable
amount of voting-eligible citizens that may be experiencing difficulty.

[^3]: MIT Election Data and Science Lab, 2017, "U.S. House 1976--2022",
    <https://doi.org/10.7910/DVN/IG0UN2>,

Our results may be of interest to key government officials, as well as
election precinct organizers. While our analysis focused on whether one
party experiences more difficulty than the other, future studies may be
interested in identifying the root causes of difficulty. This may yield
actionable insight on how the United States can restructure its voting
processes, and hopefully reduce the difficulty experienced by American
voters.


